---
title: "Project proposal"
author: "Fantastic FouR: Hannah, Eli, Preetha"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, echo = F}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      fig.align = "center")
```

```{r load-packages}
library(tidyverse)
library(broom)
library(patchwork)
library(knitr)
```

```{r load-data}
fire <- readr::read_csv('data/all_conditions.csv')
```

### Section 1. Introduction

For our project, we will be looking at which factors are most dominant in 
predicting forest fires in the state of California. Climate change represents an existential threat to humanity, and we are already seeing its effects, 
particularly on America’s west coast. We want to understand the specific factors 
that contribute to fires to more fully understand what can be done to prevent 
these tragedies from spreading. The research question we wish to look at is: 
What are the strongest environmental predictors in forest fires? We think that precipitation, humidity, Eto, and air temperature will all be strongly 
correlated with forest fires.
  


### Section 2. Data description

There are 128,125 observations in the data set. Each observation represents 
information on the weather conditions on a given weather station on a specific 
date. Each observation looks at a variety of factors such as Eto, precipitation, 
solar radiation, average vapor, maximum air temperature, minimum air 
temperature, average air temperature, maximum humidity, minimum humidity, 
average humidity, dew points, wind speed, soil temperature. The response 
variable we will be investigating is Target, which corresponds to fires on the respective observation date, in the observation region. The Target variable is a 
binary indicator, with a value of 1 indicating there was a fire and a value of 0 
indicating there was not a fire. Each of the predictors represents a different 
aspect of the climate in California. Eto looks at the presence of a natural 
flammable gas, precipitation and solar radiation look at the amount of rain, 
snow, or sunshine an area will receive on the given day. Average vapor, maximum, 
minimum and average temperature and humidity, and wind speed represents 
qualities of the air. Dew points and soil temperature look at factors that 
impact the ground. This data was scraped from CIMIS (California Irrigation 
Management Information System) weather stations by github user czaloumi using a 
selenium chromedriver. The dataset was combined with Wikipedia tables listing 
California fires by county and city to create the Target column, which indicates 
whether or not there was a fire on a particular day. Additionally, the curator 
of the dataset adds that this dataset was “used in conjunction to building an 
XGBoost Classifier to accurately predict probability for fire given 
environmental condition feature.”
  



### Section 3. Glimpse of data 
```{r}
glimpse(fire)
```


