---
title: "forcasting forest fires"
author: 'fantastic fouR: Eli Levine, Hannah Long, Preetha Ramachandran'
date: "Oct. 26, 2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

Your written report goes here! Before you submit, make sure your code chunks are turned off with `echo = FALSE` and there are no warnings or messages with `warning = FALSE` and `message = FALSE` 

```{r load-packages}
library(tidyverse)
library(broom)
library(patchwork)
library(knitr)
library(yardstick)
library(Stat2Data)
library(rms)
```

```{r load-data}
fire <- readr::read_csv('data/all_conditions.csv')
```

```{r}
fire %>%
  mutate(Target = factor(Target))
```

### Introduction

Climate change represents an existential threat to humanity. Its effects have 
the potential to dramatically shape life as we know it, creating climate refugees, 
resource wars, and submerging major cities around the globe. However, unlike 
previous challenges to our way of life, the threat of climate change will not 
manifest in a single event, rather in a series of natural disasters that will 
eventually escalate to a point where we no longer have the resources to manage 
them. This is being seen already in California as wildfires sweep the state, 
forcing people to relocate, causing issues with access and use of electricity and 
causing an estimated $10 billion in damages. Thus, in our project, we plan to 
determine what are the strongest environmental predictors of forest fires by 
looking at data from California. 

It is important to understand what these predictors are to determine how to 
prevent short term fires from escalating and to correct these conditions in the 
long run to see if it is possible to mitigate the threat of these fires going 
forward. If we know what predictors play a role in a fire, it can help authorities 
better understand what days or seasons present a higher risk, and thus prepare 
accordingly. While climate change will continue to affect our way of life, 
insights into how to manage its consequences and effects will help us plan for 
both the short term and long term future, at least until policy and research 
catch up to the severity of the issue.

Thus, the research question we want to answer is: what are the strongest 
environmental predictors of forest fires in California?

The data we will be using was scraped from CIMIS (California Irrigation 
Management Information System) weather stations by github user czaloumi using a 
selenium chromedriver. The dataset was combined with Wikipedia tables listing 
California fires by county and city to create the Target column, which indicates 
whether or not there was a fire on a particular day. Additionally, the curator 
of the dataset adds that this dataset was “used in conjunction to building an 
XGBoost Classifier to accurately predict probability for fire given environmental 
condition feature.” This user’s data contains a mixture of environmental and 
geospatial data to understand the size and the scope of the forest fires, as well 
as where the fires seem to be most frequent.

### Our Data

There are 128,126 observations in the data set. Each observation represents 
information on the weather conditions at a given weather station on a specific 
date.

The response variable we will be investigating is Target, which corresponds to 
fires on the respective observation date, in the observation region. The Target 
variable is a binary indicator, with a value of 1 indicating there was a fire 
and a value of 0 indicating there was not a fire.

Our potential predictor variables are:

**`ETo`** - The ETo variable measures the average amount of evapotranspiration 
present in the soil in each of the regions. This means that it is the amount of 
water transferred to the land by means of plants.

**`precip`** - The precip variable measurse the long term monthly average amount 
of precpitation found in the each station's region.

**`solrad`** - The solrad variable measures the average amount of solar 
radiation found in the each station's region.

**`avgvappress`** - The avgvappress variable measures the average amount of vapor 
pressure found in the each station's region.

**`avgsoiltemp`** -  The avgsoiltemp variable measures the average soil 
temperature found in the each station's region.

**`windrun`** - The windrun variable measures the sum of wind speed over a month.

**`avgwindspeed`** - The avgwindspeed variable measures the average wind speed 
found in the each station's region.

**`dewpoint`** - The dewpoint variable measures the average temperature of the 
dew on the grass in each station over a month long period.

**`avgrelhum`** - The avgrelhum variable measures the average relative humidity 
found in the each station's region.

**`avgairtemp`** - The avgairtemp variable measures the monthly average 
of the air temperature found in the each station's region.

### Exploratory Data Analysis

```{r}
newfire <- fire %>% 
  mutate(ETo = `ETo (in)`,
         precip = `Precip (in)`,
         solrad = `Sol Rad (Ly/day)`,
         avgvappress = `Avg Vap Pres (mBars)`,
         maxairtemp = `Max Air Temp (F)`,
         avgsoiltemp = `Avg Soil Temp (F)`,
         windrun = `Wind Run (miles)`,
         avgwindspeed = `Avg Wind Speed (mph)`,
         dewpoint = `Dew Point (F)`,
         avgrelhum = `Avg Rel Hum (%)`,
         minrelhum = `Min Rel Hum (%)`,
         maxrelhum = `Max Rel Hum (%)`,
         avgairtemp = `Avg Air Temp (F)`,
         minairtemp = `Min Air Temp (F)`)
```

As previously stated, the dataset contains observations of weather conditions 
and indicates the presence of a fire on a specific date at a certain weather
station in California. Each of the stations in our dataset has recorded 
observations of these weather conditions between 2018 and summer 2020. 
Because the conditions recorded by one station will likely be similar to those 
in a nearby station and similar to the recordings the day before, we had to 
simulate independence by filtering our data. 

First, observations were grouped by station id number. Next, one day was 
chosen at random from each of the stations. This gave us 153 observations 
instead of the original dataset of 128,126 observations.

By reducing the data set to a small training set, we are able to make a model 
that is a more realistic approximation of the conditions that might cause a 
fire, sans the correlation that comes with keeping all of the data. 
Additionally, constructing a model with a random sample of the data allows us to
test our final model on a new sample of data to assess its predictive power.
What more, taking a smaller, random sample of the data allows us to ensure that 
randomness is satisfied for our analysis. Finally, taking a random sample 
would ensure that the data satisfies independence, because whether a fire is 
reported or not is no longer conditional on surrounding stations, as each station 
is from different days and from different times of the year. Thus, while the 
dataset in its entirety does not satisfy independence, our random sample 
meets this criteria.

```{r}
set.seed(19)

newestfire <- newfire %>% 
  group_by(`Stn Id`) %>% 
  sample_n(1, replace = TRUE)
```

Even after reducing our dataset to 153 random observations, it was clear that
there was still some stations with missing observations in some variable 
categories. We decided to use only complete observations in the analysis, and 
thus, the total number of observations was further reduced to 143.

On surface, it did not appear that the observations with missingness differed 
systematically from the complete observations; it is thus unlikely that our 
resulting analysis is biased by the decision to remove the data.

```{r}
newestfire <- newestfire %>% 
  drop_na()
```

Our first step of our exploratory data analysis was to look at the shape of the
distributions of each variable. This would give us a sense about which 
transformations might be necessary.

```{r fig.width = 10, fig.height = 6}
p1 <- ggplot(data = newestfire, aes(x = ETo)) +
  geom_histogram(stat = "count", fill = "red") +
  labs(x = "ETo (in)",
       y = "Count",
       title = "Distribution of ETo (in)") +
  theme_minimal()

p2 <- ggplot(data = newestfire, aes(x = precip)) +
  geom_histogram(stat = "count", fill = "orange") +
  xlim(0,0.5) +
  ylim(0,10) +
  labs(x = "Precip (in)",
       y = "Count",
       title = "Distribution of Precip (in)") +
  theme_minimal()

p3 <- ggplot(data = newestfire, aes(x = solrad)) +
  geom_histogram(stat = "count", fill = "salmon") +
  labs(x = "Sol Rad (Ly/day)",
       y = "Count",
       title = "Distribution of Sol Rad (Ly/day)") +
  theme_minimal()

p4 <- ggplot(data = newestfire, aes(x = avgsoiltemp)) +
  geom_histogram(stat = "count", fill = "red") +
  labs(x = "Avg Soil Temp (F)",
       y = "Count",
       title = "Distribution of Avg Soil Temp (F)") +
  theme_minimal()

p5 <- ggplot(data = newestfire, aes(x = windrun)) +
  geom_histogram(stat = "count", fill = "orange")  +
  xlim(0,250) +
  ylim(0,1.5) +
  labs(x = "Wind Run (miles)",
       y = "Count",
       title = "Distribution of Wind Run (miles)") +
  theme_minimal()

p6 <- ggplot(data = newestfire, aes(x = avgwindspeed)) +
  geom_histogram(stat = "count", fill = "salmon") +
  labs(x = "Avg Wind Speed (mph)",
       y = "Count",
       title = "Distribution of Avg Wind Speed (mph)") +
  theme_minimal()

p7 <- ggplot(data = newestfire, aes(x = dewpoint)) +
  geom_histogram(stat = "count", fill = "red") +
  labs(x = "Dew Point (F)",
       y = "Count",
       title = "Distribution of Dew Point (F)") +
  theme_minimal()

p8 <- ggplot(data = newestfire, aes(x = avgrelhum)) +
  geom_histogram(stat = "count", fill = "orange") +
  labs(x = "Avg Relative Humidity (%)",
       y = "Count",
       title = "Distribution of Avg Relative Humidity (%)") +
  theme_minimal()


p11 <- ggplot(data = newestfire, aes(x = avgvappress)) +
  geom_histogram(stat = "count", fill = "salmon") +
  labs(x = "Avg Vap Pres (mBars)",
       y = "Count",
       title = "Distribution of Avg Vap Pres (mBars)") +
  theme_minimal()



p14 <- ggplot(data = newestfire, aes(x = avgairtemp)) +
  geom_histogram(stat = "count", fill = "red") +
  ylim(0,3) +
  labs(x = "Avg Air Temp (F)",
       y = "Count",
       title = "Distribution of Avg Air Temp (F)") +
  theme_minimal()

(p1 + p2 + p3)/(p4 + p5) 

(p6 + p7+ p14)/(p11 + p8)
```

Precipitation and average wind speed appear to be right skewed, ETo is roughly 
trimodal, average relative humidity is bimodal, and dew point is roughly left 
skewed. Across the board, the histograms are far from normal, the multiple peaks
and unique spreads and distributions are likely due to limited data points, a
hypothesis made clear by the low counts shown on each graph. 

### Methodology

To reduce multicollinearity, we decided to focus on only 
specific variables. For example, the dataset included the average, minimum and 
maximum value for a number of variables. We determined that average was likely 
the most relevant value for each condition recorded by each station throughout 
the day.

Since we are trying to predict the presence of a fire with a binary response 
variable, target, we will use a logistic regression for our analysis. 

################################################################################
However, to be able to use logistic regression we need
to check the conditions, and ensure that the data satisfies linearity, randomness
and independence.

To check linearity, we will make an empirical logistic regression plot for each 
of the predictor variables.

```{r fig.width = 7, fig.height = 3}


emplogitplot1(Target ~ ETo, data = newestfire, 
                  ngroups = 10)
emplogitplot1(Target ~ solrad, data = newestfire, 
                  ngroups = 10)
emplogitplot1(Target ~ avgvappress, data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ avgsoiltemp, data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ windrun, data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ avgwindspeed, data = newestfire, 
                  ngroups = 10)
emplogitplot1(Target ~ dewpoint, data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ avgrelhum, data = newestfire, 
                  ngroups = 10)
emplogitplot1(Target ~ avgairtemp, data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ precip, data = newestfire, 
                 ngroups = 2)

```

After looking at these graphs, it is apparent that average wind speed, and average
relative humidity do not follow a linear relationship. Furthermore, linearity is
only satisfied for precipitation when there are only 2 groups. This means that a
linear model might not be an appropriate estimation for precipitation.

#maybe move this to EDA where we actually do this process
We took a random sample of the days in our data set. This also supports 
independence. Our cleaning of the data to include a random sample of days and 
stations.
################################################################################

We then began to construct prediction models. We first started with a main
effects model containing all possible predictors.

```{r}
fire_model <- glm(Target ~ ETo + solrad + avgvappress + avgsoiltemp + windrun + 
                    avgwindspeed + avgrelhum + precip +
                    dewpoint + avgairtemp, data = newestfire, 
                  family = "binomial")
tidy(fire_model) %>%
  kable(digits = 3)
```

Because the only significant term in the model is the term for average
soil temperature (the only term with an associated p.value of less than 0.05), 
we suspected multicollinearity. We used vif to further investigate this idea.

```{r}
vif(fire_model) %>%
  kable(digits = 3)
```

We removed the variable representing the sum of wind speed over the month 
(windrun) due to its multicollinarity with average wind speed, as indicated 
by the large and similar vif values for both. Average relative humidity and 
dewpoint were removed as well due to multicollinarity with average vapor 
pressure and average air temperature, respectively.
We then constructed a new model (main_fire_model) without the aforementioned
variable.

```{r}
#removed variables with high multicollinearity: windrun, avgrelhum
main_fire_model <- glm(Target ~ ETo + solrad + avgvappress + avgsoiltemp + 
                    avgwindspeed + precip + dewpoint + 
                    avgairtemp, data = newestfire, 
                  family = "binomial")
tidy(main_fire_model) %>%
  kable(digits = 3)
```

```{r}
vif(main_fire_model) %>%
  kable(digits = 3)
```

Because the vif values are 1) dissimilar from each other or 2) generally small, 
we can conclude that we have removed highly correlated variables from analysis.

Next, using backwards selection from main_fire_model, we constructed 
new_fire_model.

```{r}
new_fire_model <- step(main_fire_model, direction = "backward")
tidy(new_fire_model, conf.int = TRUE) %>%
  kable(digts = 3)
```

```{r}
glance(new_fire_model)%>%
  select(AIC, BIC) %>%
  kable(digits = 3)
```

**write out new_fire_model here**

We then decided to try out variable transformations to potentially
bolster the predictive power of new_fire_model.

From a theoretical perspective, it is likely that our response variable, the 
log likelihood of a fire, and one of our predictors, average temperature of the 
dew on the grass (dewpoint), have a curvilinear relationship. A low value for 
dewpoint could be recorded by a particular station as a result of firefighting 
efforts while a high value for dewpoint could be the result of a fire.

To test if this might be the case, we added a quadratic
transformation of dewpoint as a predictor to our main effects 
model and fit a new model. 

```{r}
#main model, with quadratic transformation on avgwindspeed and avgrelhum
fire_model2 <- glm(Target ~ ETo + solrad + avgvappress + avgsoiltemp + windrun + 
                    precip + I(dewpoint^2) + avgairtemp + avgwindspeed + 
                     avgrelhum, 
                   data = newestfire, 
                  family = "binomial")
tidy(fire_model2) %>%
  kable(digits = 3)
```

```{r}
fire_model3 <- step(fire_model2, direction = "backward")
tidy(fire_model3, conf.int = TRUE) %>%
  kable(digits = 3)
```

```{r}
glance(fire_model3) %>%
  select(AIC, BIC) %>% 
  kable(digits = 3)
```

Adding the quadratic transformed variable, AIC has a three point improvement
over the main effects model, but BIC is larger. Because we have no preference 
for a parsimonious model (what is indicated by a lower value of BIC), we 
decided to keep the quadratic term for dewpoint. Thus, our current model is:

$$\hat{Target}  = -6.727 + 157.331ETo + -0.070solrad + 0.434avgsoiltemp + 5.255windrun + 0.002(dewpoint^2) + -0.374avgairtemp -127.386avgwindspeed$$

Next, we explored potentially meaningful interaction terms. We ultimately chose 
to test the only interaction term which seemed meaningful: ETo*avgwindspeed. We 
inferred that large amounts of water transferred to the land by means of plants 
(ETo) and high wind speed would (jointly) significantly increase the log 
likelihood of a forest fire because a lot 
of fast-moving wind might spread the water and thus make it harder for a fire to 
develop in the area, and, on the other end of the spectrum, easier for a fire to 
begin in a region with less plant based transpiration.

To determine if this interaction term is statistically significant, we added it
to the model with the quadratic dewpoint term (shown above) and conducted a 
drop-in-deviance test between the model with and without the interaction term. 

```{r}
interaction_fire_model <- glm(Target ~ ETo*avgwindspeed + solrad + avgsoiltemp + windrun + I(dewpoint^2) + 
                                avgairtemp, data = newestfire, 
                  family = "binomial")
tidy(interaction_fire_model) %>%
  kable(digits = 3)
```

```{r}
anova(fire_model3, interaction_fire_model, test = "Chisq") %>%
  tidy() %>%
  kable(digits = 3)
```

The p-value of the drop-in-deviance test is 0.761, much greater than our 
alpha level of 0.05, which suggests that the data do not provide sufficient
evidence to suggest that the interaction term is statistically significant. So, 
we will not include the interaction term in our final model.

After all these analyses/tests, our ultimate best model is:

**do we have to show it again? or is just writing it out good enough?**
```{r}
tidy(fire_model3) %>%
  kable(digits = 3)
```

$$\hat{Target}  = -6.727 + 157.331(ETo) -0.070(solrad) + 0.434(avgsoiltemp) + 5.255(windrun) + 0.002(dewpoint^2) + -0.374(avgairtemp) -127.386(avgwindspeed)$$

### Conclusion

To test our model, we first constructed an ROC curve to identify a prediction 
threshold.

```{r}
resp_aug <- augment(fire_model3) %>%
  mutate(Target = factor(Target)) %>% 
  mutate(Target = fct_relevel(Target, c("1", "0")))
```

```{r}
# calculate sensitivity and specificity at each threshold
roc_curve_data <- resp_aug %>%
  roc_curve(Target, .fitted) %>%
  mutate(false_rate = 1 - specificity) 
```
```{r}
autoplot(roc_curve_data)
```

From this ROC curve, we selected a prediction threshold of X, identified by 
minimizing 1-specificity while maximizing sensitivity. Because there is greater
risk in failing to predict a fire (type 2 error), we were less interested in the
false positive rate as opposed to the sensitivity.

```{r}
roc_curve_data %>%
  slice(1:5)
roc_curve_data %>% 
  filter(sensitivity > 0.74, sensitivity < 1, specificity < 0.8)
```

((ROC curve to identify prediction threshold))

((try model on a test dataset?))

### Discussion

Since new_fire_model appears to be our strongest model, that means that ETo, 
average air temperature and average soil temperature are the most significant 
environmental predictors of forest fires in California. While this might mean 
that monitoring these three will provide a reduction strategy to fires, it 
creates some ethical questions. ETo, ethylene oxygen, is a flammable colorless 
gas that cannot be simply removed from the atmosphere without introducing 
another agent. Furthermore, managing average temperature also provides a number 
of practical issues, as does managing soil temperature. Thus, our model suggest 
that it is likely too late to be able to do anything effectively to prevent or 
can firefighting measures only be reactive from this point forward.

The reliability and validity of our data certainly comes into question. As 
previously stated, a single data point was randomly chosen from each station 
(as the data spans multiple years and the goal was to reduce multicollinearity 
as much as possible). However, this method is not foolproof. Stations that are 
spatially close together and whose randomly selected dates are close together 
are not screened for in our data selection process. With more time, this data 
selection process would be further refined to ensure data points are as 
independent as possible.

Additionally, we only considered one potentially meaningful interaction term, 
ETo*avgwindspeed, throughout our analysis. In an expanded version of this 
project, we would potentially explore more interaction terms, as this single 
term was ultimately left out of the model. 

Though we considered both AIC and BIC throughout our analysis, we were partial 
to BIC, favoring a parsimonious model. The result is our final model with only 
three terms. With more time, we could construct 1) a model with AIC selection 
criterion and 2) a model with BIC selection criterion and compare the two on a 
new randomly selected set of data points to identify which has greater 
prediction accuracy. 

### Appendix
```{r}
p9 <- ggplot(data = newestfire, aes(x = minrelhum)) +
  geom_histogram(stat = "count") +
  labs(x = "Min Relative Humidity (%)",
       y = "Count",
       title = "Distribution of Min Relative Humidity (%)") +
  theme_minimal()

p10 <- ggplot(data = newestfire, aes(x = maxrelhum)) +
  geom_histogram(stat = "count", fill = "black") +
  labs(x = "Max Relative Humidity (%)",
       y = "Count",
       title = "Distribution of Max Relative Humidity (%)") +
  theme_minimal()

p12 <- ggplot(data = newestfire, aes(x = maxairtemp)) +
  geom_histogram(stat = "count", fill = "blue") +
  labs(x = "Max Air Temp (F)",
       y = "Count",
       title = "Distribution of Max Air Temp (F)") +
  theme_minimal()

p13 <- ggplot(data = newestfire, aes(x = minairtemp)) +
  geom_histogram(stat = "count", fill = "lightblue") +
  labs(x = "Min Air Temp (F)",
       y = "Count",
       title = "Distribution of Min Air Temp (F)") +
  theme_minimal()
(p9 + p10) /(p12 + p13)
```



------------ notes -------------------------------------------------------------
- picked variables that need transformation (rel humidity, sol rad, precip) 
      - log transform?
- build model, backward assumption


variables we want to keep:
- average humidity
multicollinearity w min and max, we think humidity prob has some effect on fire (water in air???)
- avg air temp 
multicollinearity w min and max air temp

things to do after building model w everything:
- log transform precipitation??
it looks funny
- sol radiation
closely related to soil temperature (MAKE PLOT FOR THIS FIRST)
- dew point//temperature//humidity
- wind run//wind speed


