---
title: "forcasting forest fires"
author: 'fantastic fouR: Eli Levine, Hannah Long, Preetha Ramachandran'
date: "Oct. 26, 2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r load-packages}
library(tidyverse)
library(broom)
library(patchwork)
library(knitr)
library(yardstick)
library(Stat2Data)
library(rms)
```

```{r load-data}
fire <- readr::read_csv('data/all_conditions.csv')
```

```{r, include=FALSE}
fire %>%
  mutate(Target = factor(Target))
```

### Introduction

Climate change represents an existential threat to humanity. Its effects have 
the potential to dramatically shape life as we know it by creating climate 
refugees, resource wars, and submerging major cities around the globe (Denchak, 
2017). However, unlike previous challenges to our way of life, the threat of 
climate change will not manifest in a single event, rather, in a series of 
natural disasters that will eventually escalate to a point where we no longer 
have the resources to manage these crises. 
This is being seen already in California as 
wildfires sweep the state, forcing people to relocate, causing issues with 
access and use of electricity, and causing an estimated $10 billion in damages
(Louie, 2020). 

In our project, we aim to determine the 
strongest environmental predictors of forest fires by looking at data from 
various weather stations in California. 

It is important to understand what the predictors for fires are for multiple
reasons. First, in the short run, we will be able to identify the 
conditions that increase the likelihood for forest fires. 
With this in mind, we can try to correct these conditions through policy and 
practice to see if it is at all feasible to mitigate the threat of forest 
fires in our future. Second, if we know what predictors play a role in 
predicting fires, fire fighting personnel may better understand which 
environmental or meteorological conditions present a greater fire risk and 
hopefully prepare accordingly. Thus, while climate change will continue to 
affect our way of life, insights into how to manage its immediate consequences
are evidently necessary.

Thus, the research question we wish to answer is: what are the strongest 
environmental predictors of forest fires in California?

The data we are using was scraped from CIMIS (California Irrigation 
Management Information System) weather stations by github user czaloumi using a 
selenium chromedriver. The dataset was combined with Wikipedia tables listing 
California fires by county and city to create the Target column, which indicates 
whether or not there was a fire on a particular day. Additionally, the curator 
of the dataset adds that this dataset was “used in conjunction to building an 
XGBoost Classifier to accurately predict probability for fire given 
environmental condition feature.” The data contains a mixture of 
environmental and geospatial data to understand the size and the scope of the 
forest fires, as well as where the fires seem to be most frequent.

### Our Data

There are 128,126 observations in the data set. Each observation represents 
information on the weather conditions at a given weather station on a specific 
date.

The response variable we are investigated is Target, which corresponds to 
fires on the respective observation date, in the observation region. The Target 
variable is a binary indicator, with a value of 1 indicating there was a fire 
and a value of 0 indicating there was not a fire.

Our potential predictor variables are:

**`ETo`** - The ETo variable measures the average amount of evapotranspiration 
present in the soil in each of the regions. This means that it is the amount of 
water transferred to the land by means of plants.

**`precip`** - The precip variable measures the monthly average amount of 
precpitation found in the each station's region in the days prior to the 
recording.

**`solrad`** - The solrad variable measures the average amount of solar 
radiation found in the each station's region in the prior days.

**`avgvappress`** - The avgvappress variable measures the average amount of
vapor pressure found in the each station's region in the days leading up to the 
recorded day.

**`avgsoiltemp`** -  The avgsoiltemp variable measures the average soil 
temperature found in the each station's region.

**`windrun`** - The windrun variable measures the sum of wind speed over a month
long period.

**`avgwindspeed`** - The avgwindspeed variable measures the average wind speed 
found in the each station's region.

**`dewpoint`** - The dewpoint variable measures the average temperature of the 
dew on the grass in each station over a month long period.

**`avgrelhum`** - The avgrelhum variable measures the average relative humidity 
found in the each station's region.

**`avgairtemp`** - The avgairtemp variable measures the monthly average 
of the air temperature found in the each station's region.

### Exploratory Data Analysis

```{r}
newfire <- fire %>% 
  mutate(ETo = `ETo (in)`,
         precip = `Precip (in)`,
         solrad = `Sol Rad (Ly/day)`,
         avgvappress = `Avg Vap Pres (mBars)`,
         maxairtemp = `Max Air Temp (F)`,
         avgsoiltemp = `Avg Soil Temp (F)`,
         windrun = `Wind Run (miles)`,
         avgwindspeed = `Avg Wind Speed (mph)`,
         dewpoint = `Dew Point (F)`,
         avgrelhum = `Avg Rel Hum (%)`,
         minrelhum = `Min Rel Hum (%)`,
         maxrelhum = `Max Rel Hum (%)`,
         avgairtemp = `Avg Air Temp (F)`,
         minairtemp = `Min Air Temp (F)`)
```

As previously stated, the dataset contains observations of weather conditions 
and indicates the presence of a fire on a specific date at a certain weather
station in California. Each of the stations in our dataset has recorded 
observations of these weather conditions between 2018 and summer 2020. 
Because the conditions recorded by one station will likely be similar to those 
in a nearby station and similar to the recordings the day before, we 
simulate independence by filtering our data. 

First, we grouped observations by station id number. Next, one day was
chosen at random from each of the stations. This gave us 153 observations 
instead of the original dataset of 128,126 observations.

By reducing the data set to a small training set, we were able to make a model 
that is a more realistic approximation of the conditions that might cause a 
fire, sans the heavy independence violation that comes with keeping all of the 
data. 
Additionally, constructing a model with a random sample of the data allows us to
test our final model on a separate sample of data to assess its predictive 
power.
The sampling procedule also allows us to ensure 
that randomness and independence is satisfied for our analysis. 
Whether a fire is reported or not is no longer conditional
on surrounding stations and surrounding environmental variables, as each 
station observation is from different days and from different times of the year.
Thus, while the dataset in its entirety does not satisfy independence or 
randomness, our random sample meets this criteria.

```{r}
set.seed(19)

newestfire <- newfire %>% 
  group_by(`Stn Id`) %>% 
  sample_n(1, replace = TRUE)
```

Even after we reduced our dataset to 153 random observations, it is clear that
there are still some stations with missing observations in some variable 
categories. We decided to use only complete observations in the analysis, 
and, thus, the total number of observations is further reduced to 143.

We looked at the missing observations to see if they differed at all 
systematically from the rest of the data. 

```{r}
newestfire %>% 
  filter(is.na(ETo)) %>% 
  select("Stn Name", "Target", "Stn Id", "CIMIS Region", "ETo") %>%
  kable(digits = 3)
```

It appears as though the stations with missing observations are completely
random, indicated by the varied names and station id numbers; it is therefore 
unlikely that our resulting analysis is biased by the decision to remove the 
data. We then proceeded with analysis.

```{r}
newestfire <- newestfire %>% 
  drop_na()
```

Our first step of our exploratory data analysis was to look at the distribution
of the response variable, Target, where 1 indicates a fire and 0 indicates no 
fire.

```{r, fig.height = 3, fig.wigth = 2}
newestfire1 <- newestfire %>% 
  mutate(Target = factor(Target)) %>% 
  mutate(Target = case_when(Target == 0 ~ "No Fire", Target == 1 ~ "Fire"))

area.color <- c("red", "pink")

ggplot(data = newestfire1, mapping = aes(x = Target)) +
  geom_bar(fill = area.color) +
  labs(x = "Target",
       y = "Count",
       title = "Count of Fires vs. No Fires") +
  theme_minimal(base_size = 8)
```

It's clear that our sampled data include significantly more "No Fire" 
observations as opposed to "Fire" observations. This is to be expected, as 
the forests are not on fire more often than they are. 

The next step was to look at the shape of the distributions of each potential 
predictor variable. This gave us a better understanding of our data and hinted 
at which distributions are exceptionally non-normal or otherwise grossly 
affected by our data sampling procedure. 

```{r fig.width = 10, fig.height = 6}
p1 <- ggplot(data = newestfire, aes(x = ETo)) +
  geom_histogram(fill = "red") +
  labs(x = "ETo (in)",
       y = "Count",
       title = "Distribution of ETo (in)", 
       subtitle = "Avg amount of evapotranspiration in the soil") +
  theme_minimal(base_size = 8)

p2 <- ggplot(data = newestfire, aes(x = precip)) +
  geom_histogram(fill = "orange") +
  xlim(0,0.5) +
  ylim(0,10) +
  labs(x = "Precip (in)",
       y = "Count",
       title = "Distribution of Precipitation (in)") +
  theme_minimal(base_size = 8)

p3 <- ggplot(data = newestfire, aes(x = solrad)) +
  geom_histogram(fill = "salmon") +
  labs(x = "Sol Rad (Ly/day)",
       y = "Count",
       title = "Distribution of Solar Radiation (Ly/day)") +
  theme_minimal(base_size = 8)

p4 <- ggplot(data = newestfire, aes(x = avgsoiltemp)) +
  geom_histogram(fill = "orange") +
  labs(x = "Avg Soil Temp (F)",
       y = "Count",
       title = "Distribution of Avg Soil Temperature (F)") +
  theme_minimal(base_size = 8)

p5 <- ggplot(data = newestfire, aes(x = windrun)) +
  geom_histogram(fill = "pink")  +
  xlim(0,250) +
  ylim(0,1.5) +
  labs(x = "Wind Run (miles)",
       y = "Count",
       title = "Distribution of Wind Run (miles)",
       subtitle = "Sum of wind speed over a month") +
  theme_minimal(base_size = 8)

p6 <- ggplot(data = newestfire, aes(x = avgwindspeed)) +
  geom_histogram(fill = "salmon") +
  labs(x = "Avg Wind Speed (mph)",
       y = "Count",
       title = "Distribution of Avg Wind Speed (mph)") +
  theme_minimal(base_size = 8)

p7 <- ggplot(data = newestfire, aes(x = dewpoint)) +
  geom_histogram(fill = "orange") +
  labs(x = "Dew Point (F)",
       y = "Count",
       title = "Distribution of Dew Point (F)",
       subtitle = "average temperature of dew on the grass") +
  theme_minimal(base_size = 8)

p8 <- ggplot(data = newestfire, aes(x = avgrelhum)) +
  geom_histogram(fill = "pink") +
  labs(x = "Avg Relative Humidity (%)",
       y = "Count",
       title = "Distribution of Avg Relative Humidity (%)") +
  theme_minimal(base_size = 8)

p11 <- ggplot(data = newestfire, aes(x = avgvappress)) +
  geom_histogram(fill = "salmon") +
  labs(x = "Avg Vap Pres (mBars)",
       y = "Count",
       title = "Distribution of Avg Vapor Pressure (mBars)") +
  theme_minimal(base_size = 8)

p14 <- ggplot(data = newestfire, aes(x = avgairtemp)) +
  geom_histogram(fill = "red") +
  ylim(0,3) +
  labs(x = "Avg Air Temp (F)",
       y = "Count",
       title = "Distribution of Avg Air Temperature (F)") +
  theme_minimal(base_size = 8)

(p1 + p2 + p3)/(p4 + p5) 

(p11 + p8)/(p6 + p7+ p14)
```

Some of the variables in our dataset are not shown here as they were 
not used for analysis. The plots for these variables can be found in Appendix A 
along with an explanation for their elimination. 

Precipitation and average wind speed appear to be right skewed, ETo is roughly 
trimodal, average relative humidity is bimodal, and dew point is roughly left 
skewed. Across the board, the histograms are far from normal; the multiple peaks
and unique spreads and distributions are likely due to limited data points and 
California's vast geographic diversity, a hypothesis made clear by the low 
counts shown on each graph. 

### Methodology

Our aim was to predict the presence of a fire with a binary response 
variable, Target (which indicates the presence or non presence of a fire);
therefore we used a logistic regression for our analysis. 

After identifying our regression method, we began to construct prediction 
models. We started first with a main effects model, **fire_model**, containing 
all possible predictors.

```{r}
fire_model <- glm(Target ~ ETo + solrad + avgvappress + avgsoiltemp + windrun + 
                    avgwindspeed + avgrelhum + precip +
                    dewpoint + avgairtemp, data = newestfire, 
                  family = "binomial")
tidy(fire_model) %>%
  kable(digits = 3)
```

Because the only significant terms in this model are the term for average
soil temperature and solar radiation (the only terms with an associated p.value
of less than 0.05), we suspected multicollinearity amongst variables. 
We used vif to further investigate this issue.

```{r}
vif(fire_model) %>%
  kable(digits = 3)
```

We removed the variable representing the sum of wind speed over the month 
(windrun) due to its multicollinarity with average wind speed, as indicated 
by the large and similar vif values for both. Average relative humidity and 
dewpoint were removed as well due to multicollinarity with average vapor 
pressure and average air temperature, respectively. We then constructed a new 
model,**full_fire_model**, without the aforementioned variables.

```{r}
#removed variables with high multicollinearity: windrun, avgrelhum
full_fire_model <- glm(Target ~ ETo + solrad + avgvappress + avgsoiltemp + 
                    avgwindspeed + precip + 
                    avgairtemp, data = newestfire, 
                  family = "binomial")
tidy(full_fire_model) %>%
  kable(digits = 3)
```

```{r}
vif(full_fire_model) %>%
  kable(digits = 3)
```

Because the remaining vif values are 1) dissimilar from each other or 2) 
generally small, we safely concluded that we removed all highly correlated 
variables from analysis.

Next, using backwards selection from **full_fire_model**, we constructed
**reduced_fire_model**. This was done to remove unnecessary variables and 
improve the model's predictive ability.

```{r, include = FALSE}
reduced_fire_model <- step(full_fire_model, direction = "backward")
```

```{r}
tidy(reduced_fire_model, conf.int = TRUE) %>%
  kable(digts = 3)
```

```{r}
glance(reduced_fire_model)%>%
  select(AIC, BIC) %>%
  kable(digits = 3)
```

This gave us **reduced_fire_model**:
$$log-odds(\hat{Target})  = -15.229 - 0.0118340solrad + 0.2580629avgsoiltemp$$

We then tried out various variable transformations to potentially bolster the 
predictive power of our model.

From a theoretical perspective, it is likely that our response variable, the 
log likelihood of a fire, and one of our predictors, average temperature of the 
dew on the grass (dewpoint), have a curvilinear relationship. A low value for 
dewpoint could be recorded by a particular station as a result of firefighting 
efforts while a high value for dewpoint could be the result of a fire.

To see if this hypothesis is supported by our data, we graphed the relationship
between the log odds of our response variable, Target, and dewpoint.

```{r}
target_dew <- emplogitplot1(Target ~ dewpoint, data = newestfire, 
                  ngroups = 10)
```

While not exactly quadratic, the relationship appears distinctly non-linear.

With this in mind, we tested a quadratic transformation of 
dewpoint as a predictor to our main effects model and fit a new model, 
**main_fire_model**. 

```{r}
#main model, with quadratic transformation on dewpoint
main_fire_model <- glm(Target ~ ETo + solrad + avgvappress + avgsoiltemp + windrun + 
                    precip + I(dewpoint^2) + avgairtemp + avgwindspeed + 
                     avgrelhum, 
                   data = newestfire, 
                  family = "binomial")
tidy(main_fire_model) %>%
  kable(digits = 3)
```

We then performed backwards selection on **main_fire_model** to construct
**final_fire_model**.

```{r include = FALSE}
final_fire_model <- step(main_fire_model, direction = "backward")
```

```{r}
tidy(final_fire_model, conf.int = TRUE) %>%
  kable(digits = 3)
```

```{r}
glance(final_fire_model) %>%
  select(AIC, BIC) %>% 
  kable(digits = 3)
```

Adding the quadratic transformed variable, AIC has a three point improvement
over **reduced_fire_model**, but BIC is larger. Because we have no preference 
for a parsimonious model (what is indicated by a lower value of BIC), we 
keep the quadratic term for dewpoint. Thus, our current model is 
**final_fire_model**:

$$log-odds(\hat{Target})  = -6.727 + 157.331ETo - 0.070solrad$$
$$+ 0.434avgsoiltemp + 5.255windrun + 0.002(dewpoint^2) - 0.374avgairtemp$$
$$- 127.386avgwindspeed$$

Next, we explored potentially significant interaction terms.
We ultimately chose to test the only one interaction term: ETo*avgwindspeed. We 
inferred that large amounts of water transferred to the land by plants 
(ETo) and high wind speed together would significantly decrease the log 
odds of a forest fire. We hypothesized that fast-moving wind may increase water
spread and thus make it harder for a fire to develop in a particular area. 

To determine if this interaction term is statistically significant, we added it
to the model with the quadratic dewpoint term (shown above) and conducted a 
drop-in-deviance test between the model with and without the interaction term. 

```{r}
interaction_fire_model <- glm(Target ~ ETo*avgwindspeed + solrad + avgsoiltemp +
                                windrun + I(dewpoint^2) + 
                                avgairtemp, data = newestfire, 
                  family = "binomial")
tidy(interaction_fire_model) %>%
  kable(digits = 3)
```

```{r}
anova(final_fire_model, interaction_fire_model, test = "Chisq") %>%
  tidy() %>%
  kable(digits = 3)
```

The p-value of the drop-in-deviance test is 0.761, much greater than our 
alpha level of 0.05, which suggests that the data do not provide sufficient
evidence to suggest that the interaction term is statistically significant. 
Thus, we did not include the interaction term in our final model.

After all these analyses/tests, our final model is still **final_fire_model**:
```{r}
tidy(final_fire_model) %>%
  kable(digits = 3)
```
$$log-odds(\hat{Target})  = -6.727 + 157.331ETo - 0.070solrad$$
$$+ 0.434avgsoiltemp + 5.255windrun + 0.002(dewpoint^2) - 0.374avgairtemp$$
$$- 127.386avgwindspeed$$
We can now interpret the coefficients of our model with our original research 
question in mind. 

If the priority is to identify the strongest environmental predictors 
of forest fires in California, our model indicates that the average amount of 
evapotranspiration present in the soil (ETo), average soil temperature, the sum
of wind speed over a month (windrun), and the average temperature of the dew on 
the grass (dewpoint) should all be taken into consideration. A one unit increase
in any of these predictors, to varied degrees, increases the odds of a fire
existing at a particular weather station. Conversely, our model indicates that 
a one unit increase in average air temperature, solar radiation, or average wind
speed decreases the odds of a fire existing at a particular weather station.

More specifically, our model shows that:

Holding all other variables constant, for every one inch increase
in the average amount of evapotranspiration present in the soil (ETo), the odds 
that a fire occurs increases by 2.128 (exp(157.331)). 

Holding all other variables constant, for every one Langley/day increase in 
solar radiation (can be understood as about half a Watt per square meter), the 
odds that a fire occurs increases by 0.932 (exp(-0.070)). 

Holding all other variables constant, for every one unit increase in the average 
soil temperature, the odds that a fire occurs increases by 1.543 (exp(0.434)). 

Holding all other variables constant, a one mile increase in the sum of wind 
speed over a month (windrun) increases 
the odds a fire occurs by 191.521 (exp(5.255)). 

##############################################################################
Holding all other variables constant, every one degree Farenheit squared 
increase in the average temperature of the dew on the grass (dewpoint)
increases the odds of a fire by 1.002 (exp(0.002)). 
##############################################################################

Holding all other variables constant, a one degree Fahrenheit increase in 
average air temperature increases the odds of a fire by 0.688 (exp(-0.374)). 

Holding all other variables constant, a one mile per hour increase in average 
wind speed increases the odds of a fire by 4.752949e-56 (exp(-127.386)).

### Conclusion

With a final model identified, logistic model conditions (linearity, randomness
and independence) was assessed.

To check linearity, we calculated an empirical logistic regression plot for each 
of the predictor variables. 

```{r}
par(mfrow = c(3,3))

emplogitplot1(Target ~ ETo, data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ solrad, data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ avgsoiltemp, data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ windrun, data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ avgwindspeed, data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ I(dewpoint^2), data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ avgairtemp, data = newestfire, 
                  ngroups = 5)

```

Potential violations of linearity are apparent in the
plots for solar radiation (solrad), (windrun), and average wind speed 
(avgwindspeed). However, the violations are not egregious enough to suggest that 
there is no linear relationship between the empirical logit and the predictor 
variables.

We checked randomness based on the context of the data and how the observations 
were collected. The dataset itself does not satisfy this condition, as nearby 
stations and observations on close dates are subject to a lack of 
randomness. That said, the way in which we filtered data (grouping by station id
and randomly selecting an observation) serves to satisfy this condition as 
well as the independence condition.

With all logistic regression conditions satisfied by our final model, 
went ahead with assessing its predictive power.

To test our model, we first constructed an ROC curve to identify a prediction 
threshold.

```{r}
resp_aug <- augment(final_fire_model, type.predict = "response") %>%
  mutate(Target = factor(Target)) %>% 
  mutate(Target = fct_relevel(Target, c("1", "0")))
```
```{r}
# calculate sensitivity and specificity at each threshold
roc_curve_data <- resp_aug %>%
  roc_curve(Target, .fitted) %>%
  mutate(false_rate = 1 - specificity) 
```
```{r, fig.height = 4, fig.width = 4, fig.align = "center"}
autoplot(roc_curve_data)
```

From this ROC curve, we selected a prediction threshold of 0.193, identified by 
minimizing 1-specificity while maximizing sensitivity. Because there is greater
risk in failing to predict a fire (type 2 error), we were less interested in the
false positive rate as opposed to high sensitivity.

```{r}
roc_curve_data %>% 
  filter(sensitivity > 0.75, sensitivity < 1, specificity > 0.955) %>%
  kable(digits = 3)
```
```{r}
model_aug <- resp_aug %>% 
  mutate(pred_resp = if_else(.fitted > 0.193, "fire", "no fire"))
```

```{r}
model_aug %>% 
  count(Target, pred_resp) %>%
  kable(digits = 3)
```

The confusion matrix indicates that the model correctly predicts the presence
or non presence of a fire in 135/143 cases or 94.41% of the time at a 
threshold level of 0.193.

To truly test our model, we randomly selected a new set of observations from 
our original dataset and assessed the model's predictive power on the new 
data points.

```{r include = FALSE}
set.seed(40)

newestfire_test <- newfire %>% 
  group_by(`Stn Id`) %>% 
  sample_n(1, replace = TRUE)
```

```{r include = FALSE}
newestfire_test <- newestfire_test %>% 
  drop_na()
```

```{r include = FALSE}
pred_probs = predict(final_fire_model, newestfire_test, type = "response")
pred_probs <- data_frame(pred_probs)
```

```{r include = FALSE}
model_test <- pred_probs %>% 
  mutate(pred_resp = if_else(pred_probs > 0.193, "fire", "no fire"))
```

```{r include = FALSE}
newestfire_test %>%
  filter(Target == 1) %>% 
  count()
```

The test dataset has 142 total observations, 8 of which indicate fires and 
134 of which indicate non fires.

```{r}
model_test %>% 
  count(pred_resp) %>%
  kable(digits = 3)
```

On the new, randomly selected test dataset, the confusion matrix indicates that 
the model correctly predicts the presence or non presence of a fire in 138/142 
cases or 97.18% at a threshold level of 0.193.

### Discussion

Based on our final model, ETo, solrad, avgsoiltemp, windrun, I(dewpoint^2), 
avgairtemp, avgwindspeed are the most significant environmental predictors of 
forest fires in California. 

While this might mean that monitoring these seven will provide a reduction 
strategy to fires, it raises some difficult policy questions. Each of these
respective variables are naturally occurring meteorological features. While we
can identify these conditions, not much can be done to alter these conditions
in the short run. Therefore, our findings can only help identify the conditions
that make a fire likely, but give little insight into what we can do to stop a 
forest fire. Being able to identify these conditions can help to automate the 
process of notifying the proper authorities of the odds a fire will occur at a 
particular station. However, even once they are aware of the conditions for 
forest fires as suggested by our model, it is probably too late to be able to 
do anything to effectively prevent the fire at that point, and firefighting 
measures can only be reactive from there.

The reliability and validity of our data certainly comes into question. As 
previously stated, a single data point was randomly chosen from each station 
(as the data spans multiple years and the goal was to reduce multicollinearity 
as much as possible). However, this method is not foolproof. Stations that are 
spatially close together and whose randomly selected dates are close together 
are not screened for in our data selection process. With more time, this data 
selection process would be further refined to ensure data points are as 
independent and as possible.

Looking at the confusion matrix, it is clear that the model is a stronger
predictor of no fires than fires. Where there is no fire, the model is correct
in 129/135 cases for an overall no fire prediction accuracy of 95.56%. On the
other hand, when there is a fire, the model predicts its presence in 6/8 cases
for an overall fire prediction accuracy of 75%. The stakes are higher when
there is the potential for a fire, so a prediction accuracy of 75% is 
concerning, to say the least. With more time and more data, the reason for this 
would be further investigated, and the threshold for our final model potentially
adjusted.

Additionally, we only considered one potentially meaningful interaction term, 
ETo*avgwindspeed, throughout our analysis. In an expanded version of this 
project, we would potentially explore more interaction terms, as this single 
term was ultimately left out of the model. 

Though we considered both AIC and BIC throughout our analysis, we were partial 
to AIC, with no preference for a parsimonious model. With more time, we could 
construct 1) a model with AIC as our selection criterion and 2) a model with BIC
selection criterion and compare the two on a new randomly selected set of data 
points to identify which has greater prediction accuracy. 

### References
CIMIS. (2020). CIMIS Overview. California Irrigation Management Information System. 
https://cimis.water.ca.gov/Default.aspx

Denchak, M. (2017). Global Climate Change: What You Need to Know. NRDC.Org. 
https://www.nrdc.org/stories/global-climate-change-what-you-need-know

Louie, D. (2020, October 10). Damage from California’s wildfires estimated at $10 
billion, experts say; local, state, federal cooperation needed. ABC7 San Francisco. 
https://abc7news.com/california-wildfires-cost-of-cal-fire-stanford-wildfire-research/6897462/#:%7E:text=Damage%20from%20California’s%20wildfires%20estimated%20at%20%2410%20billion%2C%20experts%20say,-KGO

Zaloumis, C. & CIMIS (2020, October). California Environmental Conditions Dataset 
(Version 2) [A collection of CIMIS recorded environmental conditions.]. 
https://www.kaggle.com/chelseazaloumis/cimis-dataset-with-fire-target


### Appendix

**FIGURE A**

```{r fig.width = 10, fig.height = 6}
p9 <- ggplot(data = newestfire, aes(x = minrelhum)) +
  geom_histogram(fill = "red") +
  labs(x = "Min Relative Humidity (%)",
       y = "Count",
       title = "Distribution of Min Relative Humidity (%)") +
  theme_minimal(base_size = 8)

p10 <- ggplot(data = newestfire, aes(x = maxrelhum)) +
  geom_histogram(fill = "orange") +
  labs(x = "Max Relative Humidity (%)",
       y = "Count",
       title = "Distribution of Max Relative Humidity (%)") +
  theme_minimal(base_size = 8)

p12 <- ggplot(data = newestfire, aes(x = maxairtemp)) +
  geom_histogram(fill = "salmon") +
  labs(x = "Max Air Temp (F)",
       y = "Count",
       title = "Distribution of Max Air Temp (F)") +
  theme_minimal(base_size = 8)

p13 <- ggplot(data = newestfire, aes(x = minairtemp)) +
  geom_histogram(fill = "red") +
  labs(x = "Min Air Temp (F)",
       y = "Count",
       title = "Distribution of Min Air Temp (F)") +
  theme_minimal(base_size = 8)

(p9 + p10) /(p12 + p13)
```

The dataset included the average, minimum and 
maximum value for a number of variables, including relative humidity and air
temperature. To reduce multicollinearity, we decided to eliminate these 
variables from our analysis. We determined that the daily averages for these 
variables were likely the most relevant value for each condition recorded by 
each station with regards to predicting fires.


