---
title: "forcasting forest fires"
author: 'fantastic fouR: Eli Levine, Hannah Long, Preetha Ramachandran'
date: "Oct. 26, 2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r load-packages}
library(tidyverse)
library(broom)
library(patchwork)
library(knitr)
library(yardstick)
library(Stat2Data)
library(rms)
```

```{r load-data}
fire <- readr::read_csv('data/all_conditions.csv')
```

```{r, include=FALSE}
fire %>%
  mutate(Target = factor(Target))
```

### Introduction

Climate change represents an existential threat to humanity. Its effects have 
the potential to dramatically shape life as we know it by creating climate 
refugees, resource wars, and submerging major cities around the globe (citation). 
However, unlike previous challenges to our way of life, the threat of climate 
change will not manifest in a single event, rather in a series of natural 
disasters that will eventually escalate to a point where we no longer have the 
resources to manage these crises. This is being seen already in California as 
wildfires sweep the state, forcing people to relocate, causing issues with 
access and use of electricity and causing an estimated $10 billion in damages
(citation). Thus, in our project, we plan to determine what are the strongest 
environmental predictors of forest fires by looking at data from California. 

It is important to understand what these predictors are for multiple reasons. 
Firstly, in the short run, we will be able to identify the conditions that cause 
fires to escalate. Thus, we can similarly correct these conditions in the 
long run to see if it is possible to mitigate the threat of these fires going 
forward. If we know what predictors play a role in a fire, it can help authorities 
better understand what days or seasons present a higher risk, and thus prepare 
accordingly. While climate change will continue to affect our way of life, 
insights into how to manage its immediate consequences and effects will help us 
plan for both the short term and long term future, at least until policy and 
research catch up to the severity of the issue.

Thus, the research question we want to answer is: what are the strongest 
environmental predictors of forest fires in California?

The data we will be using was scraped from CIMIS (California Irrigation 
Management Information System) weather stations by github user czaloumi using a 
selenium chromedriver. The dataset was combined with Wikipedia tables listing 
California fires by county and city to create the Target column, which indicates 
whether or not there was a fire on a particular day. Additionally, the curator 
of the dataset adds that this dataset was “used in conjunction to building an 
XGBoost Classifier to accurately predict probability for fire given environmental 
condition feature.” This user’s data contains a mixture of environmental and 
geospatial data to understand the size and the scope of the forest fires, as well 
as where the fires seem to be most frequent.

### Our Data

There are 128,126 observations in the data set. Each observation represents 
information on the weather conditions at a given weather station on a specific 
date.

The response variable we will be investigating is Target, which corresponds to 
fires on the respective observation date, in the observation region. The Target 
variable is a binary indicator, with a value of 1 indicating there was a fire 
and a value of 0 indicating there was not a fire.

Our potential predictor variables are:

**`ETo`** - The ETo variable measures the average amount of evapotranspiration 
present in the soil in each of the regions. This means that it is the amount of 
water transferred to the land by means of plants.

**`precip`** - The precip variable measurse the monthly average amount of 
precpitation found in the each station's region in the days prior to the recording.

**`solrad`** - The solrad variable measures the average amount of solar 
radiation found in the each station's region in the prior days.

**`avgvappress`** - The avgvappress variable measures the average amount of vapor 
pressure found in the each station's region in the days leading up to the 
recorded day.

**`avgsoiltemp`** -  The avgsoiltemp variable measures the average soil 
temperature found in the each station's region.

**`windrun`** - The windrun variable measures the sum of wind speed over a month
long period.

**`avgwindspeed`** - The avgwindspeed variable measures the average wind speed 
found in the each station's region.

**`dewpoint`** - The dewpoint variable measures the average temperature of the 
dew on the grass in each station over a month long period.

**`avgrelhum`** - The avgrelhum variable measures the average relative humidity 
found in the each station's region.

**`avgairtemp`** - The avgairtemp variable measures the monthly average 
of the air temperature found in the each station's region.

### Exploratory Data Analysis

```{r}
newfire <- fire %>% 
  mutate(ETo = `ETo (in)`,
         precip = `Precip (in)`,
         solrad = `Sol Rad (Ly/day)`,
         avgvappress = `Avg Vap Pres (mBars)`,
         maxairtemp = `Max Air Temp (F)`,
         avgsoiltemp = `Avg Soil Temp (F)`,
         windrun = `Wind Run (miles)`,
         avgwindspeed = `Avg Wind Speed (mph)`,
         dewpoint = `Dew Point (F)`,
         avgrelhum = `Avg Rel Hum (%)`,
         minrelhum = `Min Rel Hum (%)`,
         maxrelhum = `Max Rel Hum (%)`,
         avgairtemp = `Avg Air Temp (F)`,
         minairtemp = `Min Air Temp (F)`)
```

As previously stated, the dataset contains observations of weather conditions 
and indicates the presence of a fire on a specific date at a certain weather
station in California. Each of the stations in our dataset has recorded 
observations of these weather conditions between 2018 and summer 2020. 
Because the conditions recorded by one station will likely be similar to those 
in a nearby station and similar to the recordings the day before, we had to 
simulate independence by filtering our data. 

First, observations were grouped by station id number. Next, one day was 
chosen at random from each of the stations. This gave us 153 observations 
instead of the original dataset of 128,126 observations.

By reducing the data set to a small training set, we are able to make a model 
that is a more realistic approximation of the conditions that might cause a 
fire, sans the correlation that comes with keeping all of the data. 
Additionally, constructing a model with a random sample of the data allows us to
test our final model on a smaller sample of data to assess its predictive power.
What more, taking a smaller, random sample of the data allows us to ensure that 
randomness is satisfied for our analysis. 

Furthermore, taking a random sample would ensure that the data satisfies 
independence, because whether a fire is reported or not is no longer conditional
on surrounding stations and surroudning environmental variables, since each 
station is from different days and from different times of the year. Thus, while 
the dataset in its entirety does not satisfy independence, our random sample 
meets this criteria.

```{r}
set.seed(19)

newestfire <- newfire %>% 
  group_by(`Stn Id`) %>% 
  sample_n(1, replace = TRUE)
```

Even after reducing our dataset to 153 random observations, it was clear that
there was still some stations with missing observations in some variable 
categories. We decided to use only complete observations in the analysis, and 
thus, the total number of observations was further reduced to 143.

On surface, it did not appear that the observations with missingness differed 
systematically from the complete observations; it is thus unlikely that our 
resulting analysis is biased by the decision to remove the data.

```{r}
newestfire <- newestfire %>% 
  drop_na()
```

Our first step of our exploratory data analysis was to look at the shape of the
distributions of each variable. This would give us a sense about which 
transformations might be necessary.

```{r fig.width = 10, fig.height = 6}
p1 <- ggplot(data = newestfire, aes(x = ETo)) +
  geom_histogram(stat = "count", fill = "red") +
  labs(x = "ETo (in)",
       y = "Count",
       title = "Distribution of ETo (in)") +
  theme_minimal(base_size = 8)

p2 <- ggplot(data = newestfire, aes(x = precip)) +
  geom_histogram(stat = "count", fill = "orange") +
  xlim(0,0.5) +
  ylim(0,10) +
  labs(x = "Precip (in)",
       y = "Count",
       title = "Distribution of Precip (in)") +
  theme_minimal(base_size = 8)

p3 <- ggplot(data = newestfire, aes(x = solrad)) +
  geom_histogram(stat = "count", fill = "salmon") +
  labs(x = "Sol Rad (Ly/day)",
       y = "Count",
       title = "Distribution of Sol Rad (Ly/day)") +
  theme_minimal(base_size = 8)

p4 <- ggplot(data = newestfire, aes(x = avgsoiltemp)) +
  geom_histogram(stat = "count", fill = "orange") +
  labs(x = "Avg Soil Temp (F)",
       y = "Count",
       title = "Distribution of Avg Soil Temp (F)") +
  theme_minimal(base_size = 8)

p5 <- ggplot(data = newestfire, aes(x = windrun)) +
  geom_histogram(stat = "count", fill = "red")  +
  xlim(0,250) +
  ylim(0,1.5) +
  labs(x = "Wind Run (miles)",
       y = "Count",
       title = "Distribution of Wind Run (miles)") +
  theme_minimal(base_size = 8)

p6 <- ggplot(data = newestfire, aes(x = avgwindspeed)) +
  geom_histogram(stat = "count", fill = "salmon") +
  labs(x = "Avg Wind Speed (mph)",
       y = "Count",
       title = "Distribution of Avg Wind Speed (mph)") +
  theme_minimal(base_size = 8)

p7 <- ggplot(data = newestfire, aes(x = dewpoint)) +
  geom_histogram(stat = "count", fill = "orange") +
  labs(x = "Dew Point (F)",
       y = "Count",
       title = "Distribution of Dew Point (F)") +
  theme_minimal(base_size = 8)

p8 <- ggplot(data = newestfire, aes(x = avgrelhum)) +
  geom_histogram(stat = "count", fill = "orange") +
  labs(x = "Avg Relative Humidity (%)",
       y = "Count",
       title = "Distribution of Avg Relative Humidity (%)") +
  theme_minimal(base_size = 8)


p11 <- ggplot(data = newestfire, aes(x = avgvappress)) +
  geom_histogram(stat = "count", fill = "salmon") +
  labs(x = "Avg Vap Pres (mBars)",
       y = "Count",
       title = "Distribution of Avg Vap Pres (mBars)") +
  theme_minimal(base_size = 8)

p14 <- ggplot(data = newestfire, aes(x = avgairtemp)) +
  geom_histogram(stat = "count", fill = "red") +
  ylim(0,3) +
  labs(x = "Avg Air Temp (F)",
       y = "Count",
       title = "Distribution of Avg Air Temp (F)") +
  theme_minimal(base_size = 8)

(p1 + p2 + p3)/(p4 + p5) 

(p11 + p8)/(p6 + p7+ p14)
```

Some of the predictor variables in our dataset are not shown here as they were 
not used for analysis. The plots for these variables can be found in Appendix A 
along with an explanation for their elimination. 

Precipitation and average wind speed appear to be right skewed, ETo is roughly 
trimodal, average relative humidity is bimodal, and dew point is roughly left 
skewed. Across the board, the histograms are far from normal, the multiple peaks
and unique spreads and distributions are likely due to limited data points and 
California's vast geographic diversity, a hypothesis made clear by the low 
counts shown on each graph. 

### Methodology

We are trying to predict the presence of a fire with a binary response 
variable, Target (which indicates the presence or non presence of a fire);
therefore we will use a logistic regression for our analysis. 

After identifying our regression method, we began to construct prediction 
models. We began first with a main effects model containing all possible 
predictors.

```{r}
fire_model <- glm(Target ~ ETo + solrad + avgvappress + avgsoiltemp + windrun + 
                    avgwindspeed + avgrelhum + precip +
                    dewpoint + avgairtemp, data = newestfire, 
                  family = "binomial")
tidy(fire_model) %>%
  kable(digits = 3)
```

Because the only significant term in the model is the term for average
soil temperature (the only term with an associated p.value of less than 0.05), 
we suspected multicollinearity among variables. We used vif to further 
investigate this idea.

```{r}
vif(fire_model) %>%
  kable(digits = 3)
```

We removed the variable representing the sum of wind speed over the month 
(windrun) due to its multicollinarity with average wind speed, as indicated 
by the large and similar vif values for both. Average relative humidity and 
dewpoint were removed as well due to multicollinarity with average vapor 
pressure and average air temperature, respectively.We then constructed a new 
model (main_fire_model) without the aforementioned variable.

```{r}
#removed variables with high multicollinearity: windrun, avgrelhum
main_fire_model <- glm(Target ~ ETo + solrad + avgvappress + avgsoiltemp + 
                    avgwindspeed + precip + 
                    avgairtemp, data = newestfire, 
                  family = "binomial")
tidy(main_fire_model) %>%
  kable(digits = 3)
```

```{r}
vif(main_fire_model) %>%
  kable(digits = 3)
```

Because the remaining vif values are 1) dissimilar from each other or 2) 
generally small, we can conclude that we have removed highly correlated 
variables from analysis.

Next, using backwards selection from main_fire_model, we constructed 
new_fire_model. This was done to remove unnecessary variables and thus improve 
the model's predictive ability.

```{r}
new_fire_model <- step(main_fire_model, direction = "backward")
tidy(new_fire_model, conf.int = TRUE) %>%
  kable(digts = 3)
```

```{r}
glance(new_fire_model)%>%
  select(AIC, BIC) %>%
  kable(digits = 3)
```

$$\hat{Target}  = -15.229 + -0.0118340 solrad + 0.2580629avgsoiltemp $$

We then decided to try out variable transformations to potentially bolster the 
predictive power of our model.

From a theoretical perspective, it is likely that our response variable, the 
log likelihood of a fire, and one of our predictors, average temperature of the 
dew on the grass (dewpoint), have a curvilinear relationship. A low value for 
dewpoint could be recorded by a particular station as a result of firefighting 
efforts while a high value for dewpoint could be the result of a fire.

To test if this might be the case, we tested a quadratic transformation of 
dewpoint as a predictor to our main effects model and fit a new model. 

```{r}
#main model, with quadratic transformation on dewpoint
fire_model2 <- glm(Target ~ ETo + solrad + avgvappress + avgsoiltemp + windrun + 
                    precip + I(dewpoint^2) + avgairtemp + avgwindspeed + 
                     avgrelhum, 
                   data = newestfire, 
                  family = "binomial")
tidy(fire_model2) %>%
  kable(digits = 3)
```

```{r}
fire_model3 <- step(fire_model2, direction = "backward")
tidy(fire_model3, conf.int = TRUE) %>%
  kable(digits = 3)
```

```{r}
glance(fire_model3) %>%
  select(AIC, BIC) %>% 
  kable(digits = 3)
```

Adding the quadratic transformed variable, AIC has a three point improvement
over new_fire_model, but BIC is larger. Because we have no preference 
for a parsimonious model (what is indicated by a lower value of BIC), we 
decided to keep the quadratic term for dewpoint. Thus, our current model is:

$$\hat{Target}  = -6.727 + 157.331ETo + -0.070solrad + 0.434avgsoiltemp + 5.255windrun + 0.002(dewpoint^2) + -0.374avgairtemp -127.386avgwindspeed$$

Next, we explored potentially meaningful interaction terms. We ultimately chose 
to test the only interaction term which seemed meaningful: ETo*avgwindspeed. We 
inferred that large amounts of water transferred to the land by means of plants 
(ETo) and high wind speed would (jointly) significantly increase the log 
likelihood of a forest fire because a lot of fast-moving wind may spread the 
water and thus make it harder for a fire to develop in the area, and, on the 
other end of the spectrum, easier for a fire to begin in a region with less plant 
based transpiration.

To determine if this interaction term is statistically significant, we added it
to the model with the quadratic dewpoint term (shown above) and conducted a 
drop-in-deviance test between the model with and without the interaction term. 

```{r}
interaction_fire_model <- glm(Target ~ ETo*avgwindspeed + solrad + avgsoiltemp + windrun + I(dewpoint^2) + 
                                avgairtemp, data = newestfire, 
                  family = "binomial")
tidy(interaction_fire_model) %>%
  kable(digits = 3)
```

```{r}
anova(fire_model3, interaction_fire_model, test = "Chisq") %>%
  tidy() %>%
  kable(digits = 3)
```

The p-value of the drop-in-deviance test is 0.761, much greater than our 
alpha level of 0.05, which suggests that the data do not provide sufficient
evidence to suggest that the interaction term is statistically significant. 
Thus, we will not include the interaction term in our final model.

After all these analyses/tests, our final model is:
```{r}
tidy(fire_model3) %>%
  kable(digits = 3)
```

$$\hat{Target}  = -6.727 + 157.331(ETo) -0.070(solrad) + 0.434(avgsoiltemp) + 5.255(windrun) + 0.002(dewpoint^2) + -0.374(avgairtemp) -127.386(avgwindspeed)$$

### Conclusion

With a final model identified, logistic model conditions (linearity, randomness
and independence) can be assessed.

To check linearity, we calculated an empirical logistic regression plot for each 
of the predictor variables. 

```{r}
par(mfrow = c(3,3))

emplogitplot1(Target ~ ETo, data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ solrad, data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ avgsoiltemp, data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ windrun, data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ avgwindspeed, data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ I(dewpoint^2), data = newestfire, 
                  ngroups = 5)
emplogitplot1(Target ~ avgairtemp, data = newestfire, 
                  ngroups = 5)

```

Potential violations of linearity are apparent in the
plots for solar radiation (solrad), (windrun), and average wind speed 
(avgwindspeed). However, the violations are not egregious enough to suggest that 
there is no linear relationship between the empirical logit and the predictor 
variables.

We checked randomness based on the context of the data and how the observations 
were collected. The dataset itself does not satisfy this condition, as nearby 
stations and observations on between nearby dates are subject to a lack of 
randomness. That said, the way in which we filtered data (grouping by station id
and randomly selecting an observation) serves to satisfy this condition as 
well as the independence condition.

With all logistic regression conditions satisfied by our final model, 
we can go ahead with assessing its predictive power.

To test our model, we first constructed an ROC curve to identify a prediction 
threshold.

```{r}
resp_aug <- augment(fire_model3, type.predict = "response") %>%
  mutate(Target = factor(Target)) %>% 
  mutate(Target = fct_relevel(Target, c("1", "0")))
```
```{r}
# calculate sensitivity and specificity at each threshold
roc_curve_data <- resp_aug %>%
  roc_curve(Target, .fitted) %>%
  mutate(false_rate = 1 - specificity) 
```
```{r}
autoplot(roc_curve_data)
```

From this ROC curve, we selected a prediction threshold of 0.193, identified by 
minimizing 1-specificity while maximizing sensitivity. Because there is greater
risk in failing to predict a fire (type 2 error), we were less interested in the
false positive rate as opposed to high sensitivity.

```{r}
roc_curve_data %>% 
  filter(sensitivity > 0.75, sensitivity < 1, specificity > 0.95) %>%
  kable(digits = 3)
```
```{r}
model_aug <- resp_aug %>% 
  mutate(pred_resp = if_else(.fitted > 0.193, "fire", "no fire"))
```

```{r}
model_aug %>% 
  count(Target, pred_resp) %>%
  kable(digits = 3)
```
The confusion matrix indicates that the model correctly predicts the presence
or non presence of a fire in 135/143 cases or 94.41% of the time at a 
threshold level of 0.193.

To truly test our model, we randomly selected a new set of observations from 
our original dataset and assessed the model's predictive power on the new 
data points.

```{r}
set.seed(40)

newestfire_test <- newfire %>% 
  group_by(`Stn Id`) %>% 
  sample_n(1, replace = TRUE)
```

```{r}
newestfire_test <- newestfire_test %>% 
  drop_na()
```

```{r}
#main model, with quadratic transformation on dewpoint
fire_model_test <- glm(Target ~ ETo + solrad + avgsoiltemp + windrun + 
                    I(dewpoint^2) + avgairtemp + avgwindspeed, 
                   data = newestfire_test, 
                  family = "binomial")
```

```{r}
resp_test <- augment(fire_model_test, type.predict = "response") %>%
  mutate(Target = factor(Target)) %>% 
  mutate(Target = fct_relevel(Target, c("1", "0")))
```

```{r}
# calculate sensitivity and specificity at each threshold
roc_curve_data_test <- resp_test %>%
  roc_curve(Target, .fitted) %>%
  mutate(false_rate = 1 - specificity) 
```

```{r}
model_test <- resp_test %>% 
  mutate(pred_resp = if_else(.fitted > 0.193, "fire", "no fire"))
```

```{r}
model_test %>% 
  count(Target, pred_resp) %>%
  kable(digits = 3)
```
On the new, randomly selected test dataset, the confusion matrix indicates that 
the model correctly predicts the presence or non presence of a fire in 131/142 
cases or 92.25% at a threshold level of 0.193.

### Discussion

Since fire_model3 appears to be our strongest model, that means that ETo, 
solrad, avgsoiltemp, windrun, I(dewpoint^2), avgairtemp, avgwindspeed are the 
most significant environmental predictors of forest fires in California. 

While this might mean that monitoring these three will provide a reduction 
strategy to fires, it creates some difficult policy questions. Each of these
respective variables are naturally occurring features of an area. Thus while we
can identify the conditions, we likely can not do much to alter these conditions
in the short run. Therefore, our findings can only help identify the conditions
that make a fire likely, and give little insight into what we can do to stop a 
forest fire. However, being able to identify these conditions can help the 
proper authorities be made aware of the possibility that a fire is likely. 
However, our model suggest that it is likely too late to be able to do anything 
effectively to prevent or can firefighting measures only be reactive from this 
point forward.

The reliability and validity of our data certainly comes into question. As 
previously stated, a single data point was randomly chosen from each station 
(as the data spans multiple years and the goal was to reduce multicollinearity 
as much as possible). However, this method is not foolproof. Stations that are 
spatially close together and whose randomly selected dates are close together 
are not screened for in our data selection process. With more time, this data 
selection process would be further refined to ensure data points are as 
independent and as possible.

Looking at the confusion matrix, it is clear that the model is a stronger
predictor of no fires than fires. Where there is no fire, the model is correct
in 129/135 cases for an overall no fire prediction accuracy of 95.56%. On the
other hand, when there IS a fire, the model predicts its presence in 6/8 cases
for an overall fire prediction accuracy of 75%. The stakes are higher when
there is the potential for a fire, so a prediction accuracy of 75% is 
concerning, to say the least. With more time and more data, the reason for this 
would be further investigated, and the threshold for our final model potentially
adjusted.

Additionally, we only considered one potentially meaningful interaction term, 
ETo*avgwindspeed, throughout our analysis. In an expanded version of this 
project, we would potentially explore more interaction terms, as this single 
term was ultimately left out of the model. 

Though we considered both AIC and BIC throughout our analysis, we were partial 
to AIC, with no preference for a parsimonious model. With more time, we could 
construct 1) a model with AIC as our selection criterion and 2) a model with BIC
selection criterion and compare the two on a new randomly selected set of data 
points to identify which has greater prediction accuracy. 

### Appendix

**FIGURE A**

```{r}
p9 <- ggplot(data = newestfire, aes(x = minrelhum)) +
  geom_histogram(stat = "count", fill = "red") +
  labs(x = "Min Relative Humidity (%)",
       y = "Count",
       title = "Distribution of Min Relative Humidity (%)") +
  theme_minimal(base_size = 8)

p10 <- ggplot(data = newestfire, aes(x = maxrelhum)) +
  geom_histogram(stat = "count", fill = "orange") +
  labs(x = "Max Relative Humidity (%)",
       y = "Count",
       title = "Distribution of Max Relative Humidity (%)") +
  theme_minimal(base_size = 8)

p12 <- ggplot(data = newestfire, aes(x = maxairtemp)) +
  geom_histogram(stat = "count", fill = "salmon") +
  labs(x = "Max Air Temp (F)",
       y = "Count",
       title = "Distribution of Max Air Temp (F)") +
  theme_minimal(base_size = 8)

p13 <- ggplot(data = newestfire, aes(x = minairtemp)) +
  geom_histogram(stat = "count", fill = "red") +
  labs(x = "Min Air Temp (F)",
       y = "Count",
       title = "Distribution of Min Air Temp (F)") +
  theme_minimal(base_size = 8)

(p9 + p10) /(p12 + p13)
```

To reduce multicollinearity, we decided to eliminate these variables from our
analysis. For example, the dataset included the average, minimum and 
maximum value for a number of variables, including relative humidity and air
temperature. We determined that the daily averages for these variables were 
likely the most relevant value for each condition recorded by each station with
regards to predicting fires.


