---
title: "forcasting forest fires"
author: "fantastic fouR: Eli Levine, Hannah Long, Preetha Ramachandran"
date: "Oct. 26, 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

Your written report goes here! Before you submit, make sure your code chunks are turned off with `echo = FALSE` and there are no warnings or messages with `warning = FALSE` and `message = FALSE` 

```{r load-packages}
library(tidyverse)
library(broom)
library(patchwork)
library(knitr)
```

```{r load-data}
fire <- readr::read_csv('data/all_conditions.csv')
```

###Introduction

We started by creating distribution plots for each of the continuous variables. Because some of the predictor names have parentheses that aren't compatible with R, we renamed some of the variables. 

```{r}
fire <- fire %>% 
  mutate(ETo = `ETo (in)`,
         precip = `Precip (in)`,
         solrad = `Sol Rad (Ly/day)`,
         avgvappress = `Avg Vap Pres (mBars)`,
         maxairtemp = `Max Air Temp (F)`)
```

